Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-24 main.tex
Renamed files for consistency.
Minor changes.
Added bibitems

\author{
\IEEEauthorblockN{Jan Ciesko}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
jciesko@sandia.gov}
\and
\IEEEauthorblockN{David Poliakoff}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
dzpolia@sandia.gov}
\and
\IEEEauthorblockN{David S. Hollman}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Livermore, CA, USA\\
dshallm@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Center} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
}

@@

\input{src/1.Introduction}           %label: chap:introduction
\input{src/2.GenericProgramming.tex} %label: chap:background
\input{src/3.MachineModel.tex}             %label: chap:kokkosPM
\input{src/4.ProgammingModel.tex}     %label: chap:kokkosEMA
\input{src/5.Backend.tex}            %label: chap:kokkosBackend
\input{src/6.Exercises}              %label: chap:tutorial
\input{src/7.C++}                    %label: chap:c++
\input{src/8.Related}                %label: chap:related
\input{src/9.Conclusion}             %label: chap:conclusion

\begin{thebibliography}{00}
\bibitem{b1} G. Eason, B. Noble, and I. N. Sneddon, ``On certain integrals of Lipschitz-Hankel type involving products of Bessel functions,'' Phil. Trans. Roy. Soc. London, vol. A247, pp. 529--551, April 1955.
\end{thebibliography}

>>>>

\author{
\IEEEauthorblockN{Jan Ciesko}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
jciesko@sandia.gov}
\and
\IEEEauthorblockN{David Poliakoff}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
dzpolia@sandia.gov}
\and
\IEEEauthorblockN{David S. Hollman}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Livermore, CA, USA\\
dshallm@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
\and
\IEEEauthorblockN{Author Name}
\IEEEauthorblockA{\textit{Computer Science Reseach Institute} \\
\textit{Sandia National Laboratories}\\
Albuquerque, NM, USA\\
aname@sandia.gov}
}

@@

\input{src/1.Introduction}           %label: chap:introduction
\input{src/2.GenericProgramming}     %label: chap:background
\input{src/3.MachineModel}           %label: chap:kokkosPM
\input{src/4.ProgammingModel}        %label: chap:kokkosEMA
\input{src/5.Example}                %label: chap:example
\input{src/6.Tutorials}              %label: chap:tutorial
\input{src/7.C++}                    %label: chap:c++
\input{src/8.Related}                %label: chap:related
\input{src/9.Conclusion}             %label: chap:conclusion

\begin{thebibliography}{00}
\bibitem{RAJA} Hornung, Rich, Jones, Holger, Keasler, Jeff, Neely, Rob, Pearce, Olga, Hammond, Si, Trott, Christian, Lin, Paul, Vaughan, Courtenay, Cook, Jeanine, Hoekstra, Rob, Bergen, Ben, Payne, Josh, and Womeldorff, Geoff., ASC Tri-lab Co-design Level 2 Milestone Report 2015, United States, 2015
\bibitem{KOKKOS} H. Carter Edwards and Christian R. Trott and Daniel Sunderland, 
Kokkos: Enabling manycore performance portability through polymorphic memory access patterns, Journal of Parallel and Distributed Computing, Vol. 74, p. 3202-3216, 2014
\bibitem{NALU} Stefan Domino, Shreyas Ananthan, Robert Knaus, and Alan Williams, 
Deploy Nalu/Kokkos algorithmic infrastructure with performance benchmarking, ECP Milestone Report, United States, 2017
\bibitem{KOKKOSUSECASE1} Holmen, John K. and Humphrey, Alan and Sunderland, Daniel and Berzins, Martin, Improving Uintah’s Scalability Through the Use of Portable Kokkos-Based Data Parallel Tasks, Proceedings of the Practice and Experience in Advanced Research Computing 2017 on Sustainability, ACM, United States, 2017
\bibitem{KOKKOSUSECASE2} Stan Moore, LAMMPS KOKKOS Package: The quest for performance portable MD, LAMMPS Workshop, United States, 2019
\bibitem{OPENMP} https://www.openmp.org/specifications/
\bibitem{ERLANG} Cesarini, Francesco and Thompson, Simon, ERLANG Programming, O’Reilly Media, Inc., 2009
\bibitem{KOKKOS_REPO} https://github.com/kokkos/
\bibitem{KOKKOS_TUTORIAL} https://github.com/kokkos/kokkos-tutorials/
\end{thebibliography}

### 2020-01-24 src/1.Introduction.tex
Added bibitems

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS_PAPER_HERE}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL). It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{CITEKOKKOSUSECASES}. We list its key properties as follows.

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards architecture-independent and performance-portable programming of CPUs and accelerators. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

>>>>

Kokkos\footnote{from Greek for grain or seed}~\cite{KOKKOS}, is a performance-portable, parallel programming model for C++ developed at Sandia National Laboratories (SNL). It consists of a programming model specification and a library. The specification defines execution and memory model and an application programming interface (library API). The runtime library exposes programming primitives, namely, patterns and types and maps them to the underlying abstract machine model. The machine model creates a generic view of compute hardware and memories. The Kokkos library maps execution primitives to vendor-specific programming models, libraries, and memory layouts. Kokkos has been successfully used in many large-scale projects at SNL, LLNL and other research centers~\cite{KOKKOSUSECASE2}. We list its key properties as follows.

@@

Further, we would like to present Kokkos to the educational community as one representative parallel programming model for a trend towards portability abstractions. We show a set of types that implement patterns and encapsulate data with enough semantic information to produce performance-portable, parallel code on current architectures. Further, we highlight its alignment and contributions to standardization efforts towards supporting parallel programming in C++.

Secondly, we would like to promote Kokkos to students and educators as a platform that helps them to understand and experiment with parallel patterns in a fashion that is generic, vendor-agnostic and declarative. This means that students can learn about parallelism without being exposed to many architectural concerns and that they can see productivity benefits of such abstractions for performance portability when programming on current and upcoming hardware architectures.

### 2020-01-24 src/2.GenericProgramming.tex
Added bibitems

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{CITEOPENMP}, threading libraries, or languages like Erlang in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

>>>>

A programming model exposes constrained semantics that the programmer uses to express intent. Absent a programming model providing additional constraints, the semantics of C-like languages can lead to sequential thinking and execution. Parallel programming models expose parallel semantics to enable developers to create performance-portable code. This is commonly achieved by adding annotations to the base programming language, often achieved using \emph{pragma} clauses or similar language extensions. Other methods expose such semantics through library calls or new languages. All three approaches are conceptually suitable to represent programming paradigms such as tasking, data parallelism, and other patterns. Some representatives are the OpenMP programming model~\cite{OPENMP}, threading libraries, or languages like Erlang\cite{ERLANG} in which parallelism is a first-class citizen. In general, it is up to the parallel programming model implementer to expose language constructs or library interfaces (APIs) that combine convenience to the programmer with sufficiently constrained semantics to map the program execution to the parallel computer hardware efficiently and correctly. Questions of which semantic information is needed, what paradigm to present, and by what syntactic means to do so span an exciting design space that is well worth exploring.

### 2020-01-24 src/4.ProgammingModel.tex
Added bibitems

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics: Patterns (intent), Spaces, Layouts, 
Policies and Traits
Parallel Paradigms: Parallel patterns and tasking
Expression: Embedded, C++
Concern: descriptive
\end{Verbatim}
\caption{Semantic capture defined by the Kokkos programming model.}
\label{fig:SemCaptureKokkos}
\end{figure}

>>>>

\begin{figure}[h]
\begin{Verbatim}[frame=leftline]
Constrained Semantics: Patterns (intent), Spaces, 
Layouts, Policies and Traits
Parallel Paradigms: Parallel patterns and tasking
Expression: Embedded, C++
Control Paradigm: descriptive
\end{Verbatim}
\caption{Semantic capture defined by the Kokkos programming model.}
\label{fig:SemCaptureKokkos}
\end{figure}

### 2020-01-24 src/6.Tutorials.tex
Added bibitems

The Kokkos programming model distribution~\cite{CITEKOKKOSGIT} includes many examples and benchmark applications as well as tutorial and presentation material. Example applications demonstrate the use of each abstraction as well supported paradigms. Examples cover the use of unified virtual memories on CUDA devices and object polymorphism in Kokkos (virtual functions). The tutorial inclemently introduces the interested reader to Kokkos abstractions and patterns. For the purpose of education each incremental step includes two code version. One code version represents the programming assignment and the other is the solution. The assignment includes instructions and marked code sections that require completion. Benchmark applications provide further insight into performance implications and allows for easy experimentation. Further resources are online\cite{CITEKOKKOSTUTORIALS}. They include presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.

>>>>

The Kokkos programming model distribution~\cite{KOKKOS_REPO} includes many examples and benchmark applications as well as tutorial and presentation material. Example applications demonstrate the use of each abstraction as well supported paradigms. Examples cover the use of unified virtual memories on CUDA devices and object polymorphism in Kokkos (virtual functions). The tutorial inclemently introduces the interested reader to Kokkos abstractions and patterns. For the purpose of education each incremental step includes two code version. One code version represents the programming assignment and the other is the solution. The assignment includes instructions and marked code sections that require completion. Benchmark applications provide further insight into performance implications and allows for easy experimentation. Further resources are online\cite{KOKKOS_TUTORIAL}. They include presentation material as well as tutorials on application profiling and the Kokkos Kernels numerical library.
